---
title: "STA 631 Statistical Modelling And Regression Portfolio"
site: distill::distill_website
---

```{r setup }
knitr::opts_chunk$set(echo = TRUE)

```

### Loading the libraries 
```{r}
# load the libraries
library(tidyverse)
library(dplyr)
library(corrplot)
library(caret)
library(Metrics)
library(rsq)
```

### Load the Dataset
```{r }
# load the dataset
hotel_data <- read_csv("hotel_bookings.csv")
```
### Summary of the dataset 
```{r}
# overview of the dataset 

glimpse(hotel_data)

```
#### Removing missing values columns from the dataset 
```{r}
null_counts <- colSums(is.na(hotel_data))
null_counts
```


Here we can see the in children column 4 observations are missing, so we will remove the null value rows to cleanup the dataset

```{r}
hotel_clean_data <- na.omit(hotel_data)

skimr::skim(hotel_clean_data)

```
### Data Visualization 

#### 1) What is the range of years covered by the dataset, including both the earliest and latest years?

```{r}
# let's find the years the data taken from 
years_of_data <- unique(hotel_clean_data$arrival_date_year)
print(years_of_data)

```

Ans) The data is taken from 2015, 2016, 2017 approximately more than 1 lakh columns are recorded 


#### 2) What is the total number of bookings canceled and the total number of bookings not canceled? 
```{r}
# group by the  booking canceleation status 
grp_by_canceled_hotel <- hotel_clean_data %>%
  group_by(is_canceled) %>%
  summarise(count = n())

grp_by_canceled_hotel

```

Ans) From the above table we can see that there are 75166  hotel bookings are not canceled  and 44220 hotel bookings are canceled

#### 3) what percentage of bookings are canceled in the total bookings?

```{r}
# Calculate the total number of bookings
total_bookings <- sum(grp_by_canceled_hotel$count)

# Calculate the number of canceled bookings
canceled_bookings <- grp_by_canceled_hotel$count[grp_by_canceled_hotel$is_canceled == 1]

# Calculate the percentage of bookings that are canceled overall
percentage_canceled <- (sum(canceled_bookings) / total_bookings) * 100

# Print the result
cat("The percentage of bookings that are canceled overall is:", percentage_canceled, "%\n")

```
#### 4) Distribution of Canceled and Not Canceled Bookings
```{r}

# Calculate the total number of bookings
total_bookings <- sum(grp_by_canceled_hotel$count)

# Calculate the percentage of each status type
percentage <- round((grp_by_canceled_hotel$count / total_bookings) * 100, 2)

# Create labels with both status type and percentage
labels <- paste(c( "Not Canceled","Canceled"), "(", percentage, "%)", sep = " ")

# Define colors for each slice
colors <- c("green", "red")  # You can customize the colors as needed

# Plot the pie chart with labels and colors
pie(grp_by_canceled_hotel$count, labels = labels, col = colors,
    main = "Distribution of Canceled and Not Canceled Bookings")


```

#### 5) Total Number of Bookings Canceled vs. Not Canceled

```{r}
# Create a ggplot object
ggplot(hotel_clean_data, aes(x = factor(is_canceled), fill = factor(is_canceled))) +
  geom_bar() +
  labs(title = "Total Number of Bookings Canceled vs. Not Canceled",
       x = "Cancellation Status",
       y = "Total Number of Bookings") +
  scale_x_discrete(labels = c("Not Canceled", "Canceled"))

```

#### 6) What are the combined totals of bookings that were canceled and not canceled, categorized by hotel type?

```{r}
# group by the hotel type and cancelation status  
grp_by_hotel_type <- hotel_clean_data %>%
  group_by(hotel,is_canceled) %>%
  summarise(count = n())

grp_by_hotel_type
```

#### 7) Total Number of Bookings by Cancellation Status and Hotel Type

```{r}
# Create a ggplot object
ggplot(hotel_clean_data, aes(x = factor(is_canceled), fill = factor(hotel))) +
  geom_bar(position = "dodge") +
  labs(title = "Total Number of Bookings by Cancellation Status and Hotel Type",
       x = "Cancellation Status",
       y = "Total Number of Bookings") +
  scale_x_discrete(labels = c("Not Canceled", "Canceled"))


```

#### 8) What are the combined totals of bookings that were canceled and not canceled, categorized by year ?

```{r}
# group by the year and cancelation status 
grp_by_year <- hotel_clean_data %>%
  group_by(arrival_date_year,is_canceled) %>%
  summarise(count = n())

grp_by_year
```
#### 9) Total Number of Bookings by Cancellation Status over Years
```{r}
# Filter for canceled and not canceled bookings separately
canceled_bookings <- grp_by_year %>% filter(is_canceled == 1)
not_canceled_bookings <- grp_by_year %>% filter(is_canceled == 0)

# Create line plot
ggplot() +
  geom_line(data = canceled_bookings, aes(x = arrival_date_year, y = count, color = "Canceled"), size = 1) +
  geom_line(data = not_canceled_bookings, aes(x = arrival_date_year, y = count, color = "Not Canceled"), size = 1) +
  labs(
    title = "Total Number of Bookings by Cancellation Status over Years",
    x = "Year",
    y = "Total Number of Bookings") +
  scale_color_manual(values = c("red", "blue"), name = "Cancellation Status",
                     labels = c("Canceled", "Not Canceled")) +
  theme_minimal()

```

From the above graph we can see that from 2015 to 2016 cancelations and not cancelations are increased and from 2016 to 2017 cancelation and not cancelations are decreased.

#### 10) What percentage of bookings canceled in each month ?
```{r}
# Group data by 'arrival_date_month' and 'is_canceled', then calculate the total number of bookings
total_bookings <- hotel_clean_data %>%
  group_by(arrival_date_month) %>%
  summarise(total = n())

# Group data by 'arrival_date_month' and 'is_canceled', then calculate the number of canceled bookings
canceled_bookings <- hotel_clean_data %>%
  filter(is_canceled == 1) %>%
  group_by(arrival_date_month) %>%
  summarise(canceled_total = n())

# Merge total and canceled bookings dataframes
merged_data <- merge(total_bookings, canceled_bookings, by = "arrival_date_month", all.x = TRUE)

# Calculate the percentage of canceled bookings
merged_data$percentage_canceled <- (merged_data$canceled_total / merged_data$total) * 100

# Print the resulting dataframe
print(merged_data)
```
#### 11) Percentage of Bookings Canceled over Month 
```{r}

# Convert 'arrival_date_month' to factor with correct order
merged_data$arrival_date_month <- factor(merged_data$arrival_date_month, 
                                         levels = c("January", "February", "March", "April",  "May","June", "July", "August", "September", "October", "November", "December"))

# Create a ggplot object
ggplot(merged_data, aes(x = arrival_date_month, y = percentage_canceled, group = 1)) +
  geom_line(color = "blue") +
  geom_point(color = "blue", size = 3) +
  labs(
    title = "Percentage of Bookings Canceled over Month ",
    x = "Month",
    y = "Percentage Canceled") +
  scale_y_continuous(labels = scales::percent_format(scale = 1, accuracy = 1)) +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1))



```


#### 12) What are the types of rooms that were booked by customers and subsequently canceled? 

```{r}
grp_by_roomtype <- hotel_clean_data %>%
  group_by(reserved_room_type,is_canceled) %>%
  summarise(count = n())

grp_by_roomtype
```

#### 13) Number of Bookings by Reserved Room Type and Cancellation Status
```{r}
ggplot(grp_by_roomtype, aes(x = reserved_room_type, y = count, fill = factor(is_canceled))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Number of Bookings by Reserved Room Type and Cancellation Status",
       x = "Reserved Room Type",
       y = "Number of Bookings",
       fill = "Cancellation Status") +
  scale_fill_manual(values = c("Not Canceled" = "skyblue", "Canceled" = "salmon")) +
  facet_grid(. ~ factor(is_canceled)) +  
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```
#### Dropping Unwanted columns
```{r}
df <- hotel_clean_data %>%
  select(
    -arrival_date_year,
    -arrival_date_month,
    -children,
    -babies,
    -meal,
    -country,
    -market_segment,
    -distribution_channel,
    -agent,
    -company,
    -reservation_status,
    -required_car_parking_spaces,
    -reservation_status_date,
    -assigned_room_type
  )
```

### Correlation Matrix
```{r}
# Filter numeric columns
df_numeric <- df[, sapply(df, is.numeric)]

# Compute correlation matrix
correlation_matrix <- cor(df_numeric, use = "pairwise.complete.obs")

print(correlation_matrix)


```

### HeatMap 
```{r}
# Plot heatmap
corrplot(correlation_matrix, method = "color", type = "upper", 
         tl.col = "black", tl.srt = 45, tl.cex = 0.7, addCoef.col = "black", 
         number.cex = 0.7, col = colorRampPalette(c("white", "red"))(100),
         title = "Correlation Heatmap")
```



### Logistic Regression Model 

```{r}

# Convert 'is_canceled' to a factor
df$is_canceled <- as.factor(df$is_canceled)

# Set seed for reproducibility
set.seed(100)

# Create data partition
test_index <- createDataPartition(y = df$is_canceled, times = 1, p = 0.3, list = FALSE)

# Create training dataset
df_train <- df[-test_index,]

# Create validation dataset
df_validation <- df[test_index,]

# Check dimensions of training dataset
dim(df_train)
```

```{r}

# Train logistic regression model
lr_model <- glm(is_canceled ~ ., data = df_train, family = binomial)

summary (lr_model)

```

```{r}
# Make predictions on the validation dataset
lr_pred <- predict(lr_model, newdata = df_validation, type = "response")

# Convert predicted probabilities to binary predictions
lr_pred <- ifelse(lr_pred > 0.5, 1, 0)

table(df_validation$is_canceled==lr_pred)

# Print accuracy
accuracy <- sum(lr_pred == df_validation$is_canceled) / nrow(df_validation)
print(paste("Accuracy:\n", accuracy))

```

Performance Accuracy of the Logistic Model is 78 %


